Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-03-23 ThesisJP/Background/background.tex
Siste murstein på plass.

For visualization of cardiac ultrasound we want $\alpha_{\text{acc}}$ to be close to one only when a tissue interface has been encountered. If a sample of clutter noise is assigned a high opacity value, cardiac tissue further behind will not be visible. It is said to be occluded by clutter noise. The general goal in cardiac ultrasound visualization is therefore to assign high opacity to cardiac tissue and low opacity to noise samples. This is a big challenge for cardiac ultrasound since the backscattered signal level from tissue varies a lot throughout the volume. The per-volume sample value distributions for tissue and noise therefore always overlap to some degree.

@@

In \textbf{Paper IV} we extend the work by Honigmann \textit{et al.} \cite{Honigmann2003} on a global adaptive OTF by formulating an regional adaptive OTF for visualization of cardiac ultrasound volumes. This regional adaptive OTF is based on estimated statistics of the clutter noise inside the left ventricle (the blood pool). The added complexity of this technique is the statistical estimation which is basically a second volume rendering (mean and variance projection) that has to be performed before the rendered image is produced.

\section{Ultrasound field simulations}\label{sec:field}
Simulation of ultrasound probes are heavily used by researchers in order to test out new algorithms and probe designs. A simulator for ultrasound probes works by approximating the probe surface with discrete points or patches. In software it is then possible to transmit a pulse from a given probe and measure its response in space and time, the \textit{spatial impulse response}. Figure \ref{fig:hos} shows a snapshot of a simulated pressure field from an unfocused linear array emitting a continuous wave. The image was simulated with the simulator from \textbf{Paper V}. The most popular simulator for medical ultrasound field simulations is Field II \cite{Jensen1992}. All simulations in both \textbf{Paper II} and \textbf{III} are produced with this simulator. The Field II simulator works by dividing each element into mathematical elements, and all field points needs to be in the far-field of these small elements. Then it is possible to approximate the spatial impulse response by summing a trapezoid impulse response function from each mathematical element.  \todo{Bergen flops for Field II. From this we see that calculation of ultrasound simulation is complex.}

@@

Field II. Spatial impulse response.

Ultrasim. Fusk. Propose. Explain the principle behind all.

What is the computational complexity of Filed II and Ultrasim.

Simulators implemented on the GPU.

>>>>

For visualization of cardiac ultrasound we want $\alpha_{\text{acc}}$ to be close to one only when a tissue interface has been encountered. If a sample of clutter noise is assigned a high opacity value, cardiac tissue further behind will not be visible. It is said to be occluded by clutter noise. The general goal in cardiac ultrasound visualization is therefore to assign high opacity to cardiac tissue and low opacity to noise samples. This is a big challenge for cardiac ultrasound since the backscattered signal level from tissue varies a lot throughout the volume. The per-volume sample value distributions for tissue and noise therefore always overlap to some degree. With a global OTF it is therefore impossible to find a threshold that separates tissue from noise.

@@

In \textbf{Paper IV} we extend the work by Honigmann \textit{et al.} \cite{Honigmann2003} on a global adaptive OTF by formulating an regional adaptive OTF for visualization of cardiac ultrasound volumes. This regional adaptive OTF is based on estimated statistics of the clutter noise inside the left ventricle (the blood pool). The added complexity of this technique caused by the statistical estimation which is basically a second volume rendering (mean and variance projection) that has to be performed before the rendered image is produced.

\section{Ultrasound field simulations}\label{sec:field}
Simulation of ultrasound probes are heavily used by researchers in order to test out new algorithms and probe designs. A simulator for ultrasound probes typically works by approximating the probe surface with discrete points or patches. In software it is then possible to transmit a pulse from a given probe and measure its response in space and time, know as the \textit{spatial impulse response}. Figure \ref{fig:hos} shows a snapshot of a simulated pressure field from an unfocused linear array emitting a continuous wave. The image was simulated with the simulator from \textbf{Paper V}.

@@

The most popular simulator for medical ultrasound field simulations is Field II \cite{Jensen1992}. All simulations in both \textbf{Paper II} and \textbf{III} are produced with this simulator. The Field II simulator works by dividing each element into mathematical elements, and all field points needs to be in the far-field of these small elements\footnote{$l\gg\frac{w^2}{4\lambda}$, where $w$ is the mathematical element width, $l$ is the distance to the field point, and $\lambda$ is the wavelength of the transmitted pulse. For $w=\lambda/2$ the approximation is valid for $l>\lambda/16$.}. It is then possible to approximate the spatial impulse response by summing a trapezoid impulse response function originating from each mathematical element \cite{Jensen1992} in all field points. A rough estimate of the number of required FLOPS for calculating the spatial impulse respons as it is done in Field II is:
\begin{align}
\text{FLOPS}_\text{Field II} = f_s\frac{w}{c}N_{\text{elements}}N_{\text{field points}},
\end{align}
where $f_s$ is the sampling frequency, $w$ is the width of the mathematical elements, $f_s\frac{w}{c}$ is the maximal number of samples in each trapezoid function, $N_{\text{elements}}$ is the number of mathematical elements, and $N_{\text{field points}}$ is the number of field points. The required number of FLOPS for a simulation at $100$ MHz with $500$, $\lambda/2$ wide, mathematical elements across a field of $0.5$ million points is $6$ GFLOPS in order for it to complete after one second. If we want to simulate an ultrasound image the spatial impulse response on both transmit and recieve have to be calculated and convolved in time. The respons will also be different for each A-mode image. Hence, more than 800 GFLOPS is required in order to complete a Field II simulation with the given example values of an image consisting of 70 A-mode images in one second. However, Jensen and Nikolov have shown ways of reducing the computational load of Field II simulations \cite{Jensen2000}. Other simulators have also been introduced that are faster but less exact than Field II \cite{Hergum2009}.

For the simulator presented in \textbf{Paper V} the required number of FLOPS in order to achieve a certain frame rate $f_{\text{FPS}}$ is:
\begin{align}
\text{FLOPS}_{\text{HOS}} = N_{\text{source points}}N_{\text{field points}}f_{\text{FPS}}.
\end{align}
Therefore, to simulate a field of one million points with several thousand source points at 50 FPS, more than 100 GFLOPS is required. Note that this simulator evaluates a cosine and sine function for each calculation, something which makes it even harder to achieve the required throughput, since especially GPUs have less silicon devoted to such tasks.

The above examples shows that realistic and dense ultrasound simulations are computationally demanding. On the other hand we find that the levels of required processing is within the theoretical limits of modern CPUs and GPUs. Since ultrasound field simulations are quite easy to accelerate in parallel it should be possible, even if a theorethical throughput is almoste impossible to achieve in practise, to add increased interactivity to ultrasound simulations tools by utilizing parallel programing and SIMD instructions. In addition to \textbf{Paper V} threre are several simulators which have been presented the last years that does acactly this by laveraging the powers of GPUs \cite{Gjerald2012, Reichl2009, Hlawitschka2011}.
